Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-10-02 docs/papers/shallow_performance/PSEA-2015_review_response.txt

> 3. para 2, page 2: should there be 'of' on line 3? in the
  programming of models?

We're talking here about programming models such as OpenMP, CUDA etc. rather
than about constructing scientific models.

> 4. para 4, page 3: is there a version number of shallow used in the tests?

The version used in these tests has been released in version 1.0 of
the GOcean benchmark. The text has been updated accordingly.

@@

> 14. para 3, page 9: potentially change "operation of Figure 5 is
  fused" to "operation (lower loop in Figure 5) is fused."
Re-worded.

> 15. pare 1, page 10: in the sentence; "remain constant for the
> duration of each trip of the time-stepping loop" I've never heard
> that terminology before, is that standard? I would expect to hear
> something like "remain constant for the duration of each iteration
> of the time-stepping loop"
Re-worded.

> 16. the introduction of Figure 5 help to explain a bit more the
> detail of the applied transformations. This was good and it would
> have been nice to have had a bit more detail on the transformations
> from the start rather than disbursed through the text.

Unfortunately space does not permit us to describe the transformations
in more detail.

@@

> Clarification questions
> -----------------------

> Introduction

> 1) p2: In which way has the PSyKAI design been influenced by OP2? How does it compare?

Text has been added that describes how OP2 and PSyKAl compare.

@@

PSyKAl is a design approach, not a tool. Figure 1 illustrates the
structure of the Shallow code after it has been re-structured
following PSyKAl principles. How the code in the different layers
is produced is not the subject of this paper.

> Methodology

> 7) p5: Some of the transformations (3, 4, 5, 7, 9, 10) seem to be very
> specific to the Shallow application. In which way do those represent
> optimisations 'independent of the science code' as claimed?

Of necessity, the code transformations required to recover performance
for a specific code are determined by the structure of that code. We
are not claiming that a general set of optimisations can be applied to
any given code (quite the opposite in fact). Instead, the PSyKAl
approach separates the code which must be optimised from the code
describing the science. The optimisations are restricted to the PSy
layer and therefore do not affect the scientific components of the
code. This is now emphasised in the text.

@@

These recipes are implemented as python scripts that apply PSyclone
Transformations.

General comments / suggestions
------------------------------

> 1) There seems to be no preprint available yet for reference [2]

This is correct. The paper referred to has been accepted for publication
but we are awaiting notification from the publisher as to when
it will actually appear.

### 2015-10-02 docs/papers/shallow_performance/shallow_perf.tex
parco paper: final tweaks to text and response to reviewers

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal\, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal\ version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal\ version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see performance
improvement) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system. Further, the original code performed poorly using the
IBM compiler on Power8 and needed to be modified to obtain performant
code.
%%
Therefore, the \psykal\ approach can be used with negligible
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the \psykal\ version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

Achieving good performance on large numbers of light-weight cores
requires exploiting as much parallelism in an application as possible
and this results in increased complexity in the programming models
that must be used. This in turn increases the burden of code
maintenance and code development, in part because two specialisms are
required: that of the scientific domain which a code is modelling
({\it e.g.} oceanography) and that of computational science. The
situation is currently complicated still further by the existence of
competing hardware technology; if one was to begin writing a major
scientific application today it is unclear whether one would target
GPU, Xeon Phi, traditional CPU, FPGA or something else entirely. This
is a problem because, generally speaking, these different technologies
require different programming approaches.

@@

In common with OP2, the \psykal\ approach separates out the science
code and the performance related code via separate layers. The calls
that specify parallelism in both approaches are similar in terms of
where they are placed in the code and in their semantics but the
\psykal\ approach supports the specification of more than one kernel
in a parallel region of code, compared with one for OP2, giving more
scope for optimisation. Another difference is that the description
metadata is provided with a kernel in the \psykal\ approach whereas it
is provided as part of the kernel call in OP2.

@@

As the PSy layer iterates over columns, the single core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be optimised and parallelised. The PSy layer
can therefore be tailored for a particular hardware (such as
multi-core, many-core, GPGPUs, or some combination thereof) and
software (such as compiler, operating system, MPI implementation)
configuration with no change to the Algorithm or Kernel layer
code. This approach therefore offers the potential for portable
performance.

@@

For this work we use a benchmark called Shallow which solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  There is no
complete record of the people involved and the precise hardware they
were targeting. In describing our work, we term the version of the
Shallow program obtained at the beginning of this project the
`original' version.

@@

In future work we will extend our performance portability analysis of
the \psykal\ approach to shared memory parallelisation optimisations on
different architectures using different parallelisation techniques (in
particular OpenMP on multi/many-core and OpenACC on GPU's).
%
We will then analyse the performance of a domain-specific compiler
that is currently under development. This compiler will generate
optimised PSy layer code using a user-provided recipe of
transformations, thereby removing the need for optimisation experts to
manually write the PSy layer.

@@

% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
\bibliographystyle{unsrt}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{shallow_perf}
%
\end{document}

>>>>

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal\, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache-based
architectures is taken as the starting point. A vanilla \psykal\ version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal\ version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see performance
improvement) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system. Further, the original code performed poorly using the
IBM compiler on Power8 and needed to be modified to obtain performant
code.
%%
Therefore, the \psykal\ approach can be used with negligible
performance loss and sometimes small performance gains compared to the
original optimised code. We also find that there is no single best
hand-optimised implementation of the code for all of the compilers
tested.

@@

Achieving good performance on large numbers of light-weight cores
requires exploiting as much parallelism in an application as possible
and this results in increased complexity in the programming models
that must be used. This in turn increases the burden of code
maintenance and code development, in part because two specialisms are
required: that of the scientific domain which a code is modelling
({\it e.g.} oceanography) and that of computational science. The
situation is currently complicated still further by the existence of
competing hardware technology; if one were to begin writing a major
scientific application today it is unclear whether one would target
GPU, Xeon Phi, traditional CPU, FPGA or something else entirely. This
is a problem because, generally speaking, these different technologies
require different programming approaches.

@@

In common with OP2, the \psykal\ approach separates out the science
code and the performance-related code into distinct layers. The calls
that specify parallelism in both approaches are similar in terms of
where they are placed in the code and in their semantics. However, the
\psykal\ approach supports the specification of more than one kernel
in a parallel region of code, compared with one for OP2, giving more
scope for optimisation. Another difference is that the description
metadata is provided with a kernel in the \psykal\ approach whereas it
is provided as part of the kernel call in OP2.

@@

As the PSy layer iterates over columns, the single core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be exploited and optimised. The PSy layer
can therefore be tailored for a particular hardware (such as
multi-core, many-core, GPGPUs, or some combination thereof) and
software (such as compiler, operating system, MPI implementation)
configuration with no change to the Algorithm or Kernel layer
code. This approach therefore offers the potential for portable
performance.

@@

For this work we use a benchmark called Shallow which solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  There is no
complete record of the people involved and the precise hardware they
were targeting. In describing our work, we term the version of the
Shallow program obtained at the beginning of this project the
`original' version. This code is available as part of version 1.0 of
the GOcean benchmark (\url{https://puma.nerc.ac.uk/trac/GOcean}).

@@

In future work we will extend our performance portability analysis of
the \psykal\ approach to shared memory parallelisation optimisations on
different architectures using different parallelisation techniques (in
particular OpenMP on multi/many-core and OpenACC on GPU's).
%
We will then analyse the performance of a domain-specific compiler
that is currently under development. This compiler will generate
optimised PSy layer code by following a user-provided recipe of
transformations, thereby removing the need for optimisation experts to
manually write the PSy layer.

@@

\bibliographystyle{unsrt}
\bibliography{shallow_perf}
%
\end{document}
